---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

sim_formula <- bf(Distance ~ Sem + Phon + Sem*Phon)

get_prior(data = sim, sim_formula)

my_priors <- c(prior(normal(107, 50), class = 'Intercept'),
               prior(normal(0, 50), class = 'b', coef = 'SemSimilar'),
               prior(normal(0, 50), class = 'b', coef = 'PhonSimilar'),
               prior(normal(0, 50), class = 'b', coef = 'SemSimilar:PhonSimilar'))


```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

xmdl_sim <- brm(formula = sim_formula,
                 prior = my_priors,
                 family = gaussian(),
                 data = sim, 
                 chains = 4,
                 cores = 4, 
                 iter = 2000,
                 warmup = 1000,
                 thin = 1)

xmdl_sim

#YOUR MODEL HERE

```

## (c) Posterior predictive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

xmdl_sim %>% 
pp_check(ndraws = 100)

#Not perfect, Let's try the square root

sim <- sim |> 
  mutate(sqrt_dist = sqrt(Distance))

my_priors <- c(prior(normal(10, 7), class = 'Intercept'),
               prior(normal(0, 7), class = 'b', coef = 'SemSimilar'),
               prior(normal(0, 7), class = 'b', coef = 'PhonSimilar'),
               prior(normal(0, 7), class = 'b', coef = `SemSimilar:PhonSimilar`))

xmdl_sqrt <- brm(sqrt_dist ~ Sem + Phon + Sem*Phon,
            family = gaussian(),
            prior = my_priors,
            data = sim,
            chains = 4,
            cores = 4, 
            iter = 2000,
            warmup = 1000,
            thin = 1)

pp_check(xmdl_sqrt, ndraws = 100)

#Beautiful

```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics
xmdl_sqrt
mcmc_plot(xmdl_sqrt, type = 'trace')
mcmc_dens_overlay(xmdl_sqrt)

# Rhat equals 1, both Bulk_ESS and Tail_ESS are more than 2000, which is a good number

```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

#Visualization
xmdl_sqrt |> 
  spread_draws(b_SemSimilar) |> 
  ggplot() +
    stat_halfeye(aes(x = b_SemSimilar), fill = 'cyan') +
    geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Distance",
        y = "") +
   theme_minimal()

xmdl_sqrt |> 
  spread_draws(b_PhonSimilar) |> 
  ggplot() +
    stat_halfeye(aes(x = b_PhonSimilar), fill = 'cyan') +
    geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Distance",
        y = "") +
   theme_minimal()

xmdl_sqrt |> 
  spread_draws(`b_SemSimilar:PhonSimilar`) |> 
  ggplot() +
    stat_halfeye(aes(x = `b_SemSimilar:PhonSimilar`), fill = 'cyan') +
    geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Distance",
        y = "") +
   theme_minimal()
```

```{r}
#Probability of direction
xmdl_sqrt |> 
  spread_draws(b_SemSimilar) |> 
  select(b_SemSimilar) |>
  p_direction()

xmdl_sqrt |> 
  spread_draws(b_PhonSimilar) |>
  select(b_PhonSimilar) |>
  p_direction()

xmdl_sqrt |> 
  spread_draws(`b_SemSimilar:PhonSimilar`) |>
  select(`b_SemSimilar:PhonSimilar`) |>
  p_direction()

```

```{r}
#Get Posteriors
xmdl_sqrt_prior<- brm(sqrt_dist ~ Sem + Phon + Sem*Phon,
                  family = gaussian(),
                  prior = my_priors,
                  sample_prior = "only",
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1)
```

```{r}
#bayes factor
posteriors_Sem <- xmdl_sqrt |> 
  spread_draws(b_SemSimilar) |> 
  select(b_SemSimilar)

priors_Sem <- xmdl_sqrt_prior |> 
  spread_draws(b_SemSimilar) |> 
  select(b_SemSimilar)

bayesfactor_parameters(posteriors_Sem, priors_Sem, direction = "two-sided", null = 0)

posteriors_Phon <- xmdl_sqrt |> 
  spread_draws(b_PhonSimilar) |> 
  select(b_PhonSimilar)

priors_Phon <- xmdl_sqrt_prior |> 
  spread_draws(b_PhonSimilar) |> 
  select(b_PhonSimilar)

bayesfactor_parameters(posteriors_Phon, priors_Phon, direction = "two-sided", null = 0)


posteriors_Sem_Phon <- xmdl_sqrt |> 
  spread_draws(`b_SemSimilar:PhonSimilar`) |> 
  select(`b_SemSimilar:PhonSimilar`)

priors_Sem_Phon <- xmdl_sqrt_prior |> 
  spread_draws(`b_SemSimilar:PhonSimilar`) |> 
  select(`b_SemSimilar:PhonSimilar`)

bayesfactor_parameters(posteriors_Sem_Phon, priors_Sem_Phon, direction = "two-sided", null = 0)
```

## (g) Write up your entire analysis:

YOUR WRITE UP HERE

#I'm still struggling with this part I'm just going to give my best shot.

For three factors considered, the semantic similarity, phonetic similarity, and the interaction of the two, the visualization showed us that mostly the semantic similarity leads to shorter distance while phonetic similarity leads to longer distance, and the interaction leads more to longer distance rather than shorter. Based on pd analysis, the pd for both semantic similarity and and phonetic similarity are 94% (means mostly unidirectional), but pd for the interaction is only 65%. And lastly, the bayes factor for all three factors are less than 1, which means the alternative hypothesis has a larger probability to be true and thus can not be rejected.

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
